---
title: "Simple Model Review"
author: "Dina Dinh"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(easypackages)
libraries("dplyr", "tidyr", "ggplot2", "ggeffects", "mmrm", "emmeans", "knitr", "kableExtra")
```


# Comparing Model 2 and Model 3

**The baselines are unequal in each of these scenarios.**

```{r}
set.seed(1)

n_timepoint = 50
n_group = n_timepoint*2

# In this dataset, model2 and model3 do agree

df1_wide_agree <- data.frame(
  id = rep(1:n_timepoint, 2),
  group = rep(c("A", "B"), each = n_timepoint)
) %>%
  mutate(
    # Group A baseline mean = 165
    # Group B baseline mean = 150
    Baseline = ifelse(
      group == "A",
      (rnorm(n_timepoint, mean = 0, sd = 1)*15)+15+150,
      (rnorm(n_timepoint, mean = 0, sd = 1)*15)+150
    ),
    
    # Final = that person's baseline + random change
    Final = Baseline + rnorm(n(), mean = 0, sd = 1)*5,
    
    change = Final - Baseline
  )


df1_agree <- df1_wide_agree %>%
  pivot_longer(
    cols = c(Baseline, Final),
    names_to = "timepoint",
    values_to = "weight"
  )
```

```{r}
set.seed(1)

n_timepoint = 50
n_group = n_timepoint*2

# In this dataset, model2 and model3 do NOT agree

df1_wide_disagree <- data.frame(
  id = rep(1:n_timepoint, 2),
  group = rep(c("A", "B"), each = n_timepoint)
) %>%
  mutate(
    # Group A baseline mean = 165
    # Group B baseline mean = 150
    Baseline = ifelse(
      group == "A",
      rnorm(n_timepoint, mean = 70, sd = 6),
      rnorm(n_timepoint, mean = 70, sd = 6)
    ),
    
    # Final = that person's baseline + random change
    Final = ifelse(
      group == "A",
      rnorm(n(), mean = 85, sd = 6),
      rnorm(n_timepoint, mean = 72, sd = 6)
    ),
    
    change = Final - Baseline
  )


df1_disagree <- df1_wide_disagree %>%
  pivot_longer(
    cols = c(Baseline, Final),
    names_to = "timepoint",
    values_to = "weight"
  )
```

## Where Model 2 and Model 3 Agree in Contrasts

In this scenario, we see that there's a big difference at final between the two groups, BUT the within-group change is almost zero for both groups.

- Means at final are different, but change scores are equal.

```{r}
model1_agree = lm(Final ~ group, data = df1_wide_agree) 
model2_agree = lm(Final ~ Baseline + group, data = df1_wide_agree) # "group" hypothesis is same as model4 in estimate table.
model3_agree = lm(change ~ group, data = df1_wide_agree) 
model4_agree = lm(change ~ Baseline + group, data = df1_wide_agree) 

lsmeans1_agree = lsmeans(model1_agree, ~ group, adjust = "none")
lsmeans1_summary_agree = summary(lsmeans1_agree, infer = c(TRUE, TRUE), adjust = "none")
contrast_1_agree = as.data.frame(contrast(lsmeans1_agree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans2_agree = lsmeans(model2_agree, ~ group, adjust = "none")
lsmeans2_summary_agree = summary(lsmeans2_agree, infer = c(TRUE, TRUE), adjust = "none")
contrast_2_agree = as.data.frame(contrast(lsmeans2_agree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans3_agree = lsmeans(model3_agree, ~ group, adjust = "none")
lsmeans3_summary_agree = summary(lsmeans3_agree, infer = c(TRUE, TRUE), adjust = "none")
contrast_3_agree = as.data.frame(contrast(lsmeans3_agree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans4_agree = lsmeans(model4_agree, ~ group, adjust = "none")
lsmeans4_summary_agree = summary(lsmeans4_agree, infer = c(TRUE, TRUE), adjust = "none")
contrast_4_agree = as.data.frame(contrast(lsmeans4_agree, method = "pairwise", infer = c(TRUE, TRUE)))
```

```{r}
# Linear Mixed Model
df2_agree = df1_agree %>%
  group_by(group, timepoint) %>%
  mutate(id = row_number()) %>%
  ungroup() %>%
  mutate(group = as.factor(group),
         timepoint = as.factor(timepoint),
         id = ifelse(group == "B", id*100, id),
         id = as.character(id))

library(mmrm)
lmm_model1_agree = mmrm(weight ~ group * timepoint + us(timepoint | id), data = df2_agree)

lmm_lsmeans1_agree = lsmeans(lmm_model1_agree, pairwise ~ group * timepoint, adjust = "none")

contrast1_agree <- list(
"Diff of Diff" =
c(1, -1, -1, 1)
)

contrast_df1_agree = as.data.frame(lmm_lsmeans1_agree$contrasts)
contrast_df1_agree = contrast_df1_agree %>%
filter(contrast %in% c("A Baseline - B Baseline",
                       "A Baseline - A Final",
                       "B Baseline - B Final"))

contrast1.1_agree <- contrast(lmm_lsmeans1_agree, method = contrast1_agree, adjust = "none")
contrast_df1.1_agree = as.data.frame(contrast1.1_agree)
contrast_df1_agree = rbind(contrast_df1_agree, contrast_df1.1_agree)
```

```{r}
contrast_1_agree$model_type = "Model 1"
contrast_2_agree$model_type = "Model 2"
contrast_3_agree$model_type = "Model 3"
contrast_4_agree$model_type = "Model 4"
contrast_df1_agree$model_type = "LMM"

summary_contrast_agree = rbind(contrast_1_agree, contrast_2_agree, contrast_3_agree, contrast_4_agree) %>%
  dplyr::select(-lower.CL, -upper.CL)

summary_contrast_agree = rbind(contrast_df1_agree, summary_contrast_agree)
```

```{r}
p_col <- "p.value"

summary_contrast_agree %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 4)),
    across(
      all_of(p_col),
      ~ case_when(
        .x < 0.001 ~ cell_spec("<0.001", bold = TRUE),
        .x < 0.05  ~ cell_spec(as.character(.x), bold = TRUE),
        TRUE       ~ as.character(.x)
      )
    )
  ) %>%
  kable(
    format = "html",
    escape = FALSE,
    caption = "Type III Analysis of Variance Table with Satterthwaite's method"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  kableExtra::kable_styling() %>%
  kableExtra::footnote(
    general = "Table For Model 2 and Model 3 Agree"
  )
```

## Where Model 2 and Model 3 Disgree in Contrasts

In this scenario, we see that at final, there's not much change between the two groups BUT the within-group change for Group B is large.

- Means at final are same, but change scores are different.

```{r}
model1_disagree = lm(Final ~ group, data = df1_wide_disagree) 
model2_disagree = lm(Final ~ Baseline + group, data = df1_wide_disagree) # "group" hypothesis is same as model4 in estimate table.
model3_disagree = lm(change ~ group, data = df1_wide_disagree) 
model4_disagree = lm(change ~ Baseline + group, data = df1_wide_disagree) 

lsmeans1_disagree = lsmeans(model1_disagree, ~ group, adjust = "none")
lsmeans1_summary_disagree = summary(lsmeans1_disagree, infer = c(TRUE, TRUE), adjust = "none")
contrast_1_disagree = as.data.frame(contrast(lsmeans1_disagree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans2_disagree = lsmeans(model2_disagree, ~ group, adjust = "none")
lsmeans2_summary_disagree = summary(lsmeans2_disagree, infer = c(TRUE, TRUE), adjust = "none")
contrast_2_disagree = as.data.frame(contrast(lsmeans2_disagree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans3_disagree = lsmeans(model3_disagree, ~ group, adjust = "none")
lsmeans3_summary_disagree = summary(lsmeans3_disagree, infer = c(TRUE, TRUE), adjust = "none")
contrast_3_disagree = as.data.frame(contrast(lsmeans3_disagree, method = "pairwise", infer = c(TRUE, TRUE)))
lsmeans4_disagree = lsmeans(model4_disagree, ~ group, adjust = "none")
lsmeans4_summary_disagree = summary(lsmeans4_disagree, infer = c(TRUE, TRUE), adjust = "none")
contrast_4_disagree = as.data.frame(contrast(lsmeans4_disagree, method = "pairwise", infer = c(TRUE, TRUE)))
```

```{r, eval = F}
summary(model2_disagree)
summary(model3_disagree)
summary(model4_disagree)

mean(df1_disagree$weight[df1_disagree$timepoint=="Baseline" & df1_disagree$group == "A"])
mean(df1_disagree$weight[df1_disagree$timepoint=="Final" & df1_disagree$group == "A"])
mean(df1_disagree$weight[df1_disagree$timepoint=="Baseline" & df1_disagree$group == "B"])

mean(df1_disagree$weight[df1_disagree$timepoint=="Final" & df1_disagree$group == "B"])
```

```{r}
cor(df1_disagree$weight[df1_disagree$timepoint=="Baseline" & df1_disagree$group == "A"], df1_disagree$weight[df1_disagree$timepoint=="Final" & df1_disagree$group == "A"])

cor(df1_disagree$weight[df1_disagree$timepoint=="Baseline" & df1_disagree$group == "B"], df1_disagree$weight[df1_disagree$timepoint=="Final" & df1_disagree$group == "B"])
```


```{r}
# Linear Mixed Model
df2_disagree = df1_disagree %>%
  group_by(group, timepoint) %>%
  mutate(id = row_number()) %>%
  ungroup() %>%
  mutate(group = as.factor(group),
         timepoint = as.factor(timepoint),
         id = ifelse(group == "B", id*100, id),
         id = as.character(id)) 

df2_disagree_a = df2_disagree %>%
  mutate(id = as.numeric(id)) %>%
  filter(group == "A" & id <= 25)

df2_disagree_b = df2_disagree %>%
  mutate(id = as.numeric(id)) %>%
  filter(group == "B")

df2_disagree = rbind(df2_disagree_a, df2_disagree_b) %>%
  mutate(id = as.factor(id))

library(mmrm)
lmm_model1_disagree = mmrm(weight ~ group * timepoint + us(timepoint | id), data = df2_disagree)

lmm_lsmeans1_disagree = lsmeans(lmm_model1_disagree, pairwise ~ group * timepoint, adjust = "none")

contrast1_disagree <- list(
  "Diff of Diff" =
    c(1, -1, -1, 1)
)
contrast2_disagree <- list(
  "Test Contrast" =
    c(-0.5, 0.5, 1, -1)
)

contrast_df1_disagree = as.data.frame(lmm_lsmeans1_disagree$contrasts)
contrast_df1_disagree = contrast_df1_disagree %>%
  filter(contrast %in% c("A Baseline - B Baseline",
                         "A Baseline - A Final",
                         "B Baseline - B Final",
                         "A Final - B Final"))

contrast1.1_disagree <- contrast(lmm_lsmeans1_disagree, method = contrast1_disagree, adjust = "none")
contrast1.2_disagree <- contrast(lmm_lsmeans1_disagree, method = contrast2_disagree, adjust = "none")

contrast_df1.1_disagree = as.data.frame(contrast1.1_disagree)
contrast_df1.2_disagree = as.data.frame(contrast1.2_disagree)

contrast_df1_disagree = rbind(contrast_df1_disagree, contrast_df1.1_disagree, contrast_df1.2_disagree)
```

```{r}
contrast_1_disagree$model_type = "Model 1"
contrast_2_disagree$model_type = "Model 2"
contrast_3_disagree$model_type = "Model 3"
contrast_4_disagree$model_type = "Model 4"
contrast_df1_disagree$model_type = "LMM"

summary_contrast_disagree = rbind(contrast_1_disagree, contrast_2_disagree, contrast_3_disagree, contrast_4_disagree) %>%
  dplyr::select(-lower.CL, -upper.CL)

summary_contrast_disagree = rbind(contrast_df1_disagree, summary_contrast_disagree)
```

```{r}
p_col <- "p.value"

summary_contrast_disagree %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 4)),
    across(
      all_of(p_col),
      ~ case_when(
        .x < 0.001 ~ cell_spec("<0.001", bold = TRUE),
        .x < 0.05  ~ cell_spec(as.character(.x), bold = TRUE),
        TRUE       ~ as.character(.x)
      )
    )
  ) %>%
  kable(
    format = "html",
    escape = FALSE,
    caption = "Type III Analysis of Variance Table with Satterthwaite's method",
    
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  kableExtra::kable_styling() %>%
  kableExtra::footnote(
    general = "Table For Model 2 and Model 3 Disagree"
  )
```


```{r stop-running, include=FALSE}
knitr::opts_chunk$set(eval = F)
```

```{r}
set.seed(1)

n_timepoint = 50
n_group = n_timepoint*2

# In this dataset, model2 and model3 do agree

df1_wide <- data.frame(
  id = rep(1:n_timepoint, 2),
  group = rep(c("A", "B"), each = n_timepoint)
) %>%
  mutate(
    # Group A baseline mean = 165
    # Group B baseline mean = 150
    Baseline = ifelse(
      group == "A",
      (rnorm(n_timepoint, mean = 0, sd = 1)*15)+15+150,
      (rnorm(n_timepoint, mean = 0, sd = 1)*15)+150
    ),
    
    # Final = that person's baseline + random change
    Final = Baseline + rnorm(n(), mean = 0, sd = 1)*5,
    
    change = Final - Baseline
  )


df1 <- df1_wide %>%
  pivot_longer(
    cols = c(Baseline, Final),
    names_to = "timepoint",
    values_to = "weight"
  )
```

```{r, eval=FALSE}

# In this dataset, model2 and model3 do NOT agree

df1_wide <- data.frame(
  id = rep(1:n_timepoint, 2),
  group = rep(c("A", "B"), each = n_timepoint)
) %>%
  mutate(
    # Group A baseline mean = 165
    # Group B baseline mean = 150
    Baseline = ifelse(
      group == "A",
      rnorm(n_timepoint, mean = 70, sd = 6),
      rnorm(n_timepoint, mean = 80, sd = 6)
    ),
    
    # Final = that person's baseline + random change
    Final = rnorm(n(), mean = 85, sd = 6),
    
    change = Final - Baseline
  )


df1 <- df1_wide %>%
  pivot_longer(
    cols = c(Baseline, Final),
    names_to = "timepoint",
    values_to = "weight"
  )
```


```{r}
library(ggplot2)

ggplot(df1, aes(x = timepoint, y = weight, fill = group)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  coord_cartesian(ylim = c(65, 80))
  theme_minimal()
```


```{r}
# Simple T-Test

t.test(df1_wide_disagree$Baseline[df1_wide_disagree$group == "A"], df1_wide_disagree$Final[df1_wide_disagree$group == "A"], paired = T)

t.test(df1_wide_disagree$change[df1_wide_disagree$group == "A"], mu = 0)
# significant but doesn't take into consideration the timepoints.
```

```{r}
# T-Test of the differences
t.test(df1_wide$change ~ df1_wide$group)

# significant 

```

Assumptions for t-test: 

1. Independent subjects across groups
- One person in group A should not be correlated with a person in group B.

2. Change scores are approximately normally distributed within each group
- This does not mean the original outcome at time1 or time2 must each be normal. It is the change score that matters most.

3. Equal variances only if you use the pooled t-test
- If you use the default Welch t-test in R, equal variances are not required.

4. Outcome is continuous
- The change score should be roughly quantitative/continuous.

5. No major outliers in the change scores
- Extreme outliers can distort the mean and the t-test.

```{r}
#  Paired T-Test

# Group A
t.test(df1_wide$Baseline[df1_wide$group == "A"], df1_wide$Final[df1_wide$group == "A"], paired = T)
# Not significant

# Group B
t.test(df1_wide$Baseline[df1_wide$group == "B"], df1_wide$Final[df1_wide$group == "B"], paired = T)
# Significant

# these results does not really compare the differences of the timepoints between the 2 groups.
```

```{r}
# Linear Mixed Model
df2 = df1 %>%
  group_by(group, timepoint) %>%
  mutate(id = row_number()) %>%
  ungroup() %>%
  mutate(group = as.factor(group),
         timepoint = as.factor(timepoint),
         id = ifelse(group == "B", id*100, id),
         id = as.character(id))

library(mmrm)
lmm_model1 = mmrm(weight ~ group * timepoint + us(timepoint | id), data = df2)

lsmeans1 = lsmeans(lmm_model1, pairwise ~ group * timepoint, adjust = "none")

contrast1 <- list(
"Diff of Diff" =
c(1, -1, -1, 1)
)

contrast_df1 = as.data.frame(lsmeans1$contrasts)
contrast_df1 = contrast_df1 %>%
filter(contrast %in% c("A Baseline - B Baseline",
                       "A Baseline - A Final",
                       "B Baseline - B Final"))

contrast1.1 <- contrast(lsmeans1, method = contrast1, adjust = "none")
contrast_df1.1 = as.data.frame(contrast1.1)
contrast_df1 = rbind(contrast_df1, contrast_df1.1)
contrast_df1
```

Assumptions for LMM: 

1. Correct mean structure
- The model for the mean should be appropriate:
  - group effect
  - time effect
  - group × time interaction
  
2. Independence between subjects
- Different subjects should be independent.

3. Correlation within subjects is handled correctly enough
- The mixed model accounts for repeated observations within subject through the random effect or covariance structure.

4. Residuals are approximately normal
- More specifically, the residuals and random effects are often assumed approximately normal for standard inference in linear mixed models.

5. Homoscedasticity of residuals
- Residual variance should be reasonably similar across fitted values/groups/times unless you model different variances.

6. Outcome is continuous
- For standard linear mixed models.

7. No extreme influential outliers

```{r}
shapiro.test(resid(model1))
```


```{r, eval=FALSE}
mean(df1$weight[df1$timepoint=="Baseline" & df1$group == "A"])
mean(df1$weight[df1$timepoint=="Baseline" & df1$group == "B"])
mean(df1$weight[df1$timepoint=="Final" & df1$group == "A"])
mean(df1$weight[df1$timepoint=="Final" & df1$group == "B"])

mean(df1$weight[df1$group == "A"])
mean(df1$weight[df1$group == "B"])
```

# ANCOVA

At the final timepoint, do treatment and control differ after adjusting for baseline difference? -> model2 answers this

```{r}
model1 = lm(Final ~ group, data = df1_wide)
model2 = lm(Final ~ Baseline + group, data = df1_wide) # "group" hypothesis is same as model4 in estimate table.
model3 = lm(change ~ group, data = df1_wide) 
model4 = lm(change ~ Baseline + group, data = df1_wide) 
model5 = lm(Final ~ 0 + group + Baseline:group, data = df1_wide)

AIC(lmm_model1)
AIC(model2)
AIC(model3)
AIC(model4)
AIC(model5)

summary(model2)$coef
summary(model3)$coef
summary(model4)$coef
summary(model5)$coef
```

What is the relationship among the models 2-4? What do they mean and what are they assuming? For models 2-4, what is the null hypothesis for the group assuming? Which models 1-4 should we use and in what cases?

model1 : answers if there the a difference between the two groups at final
  - null hypothesis: the two group's means are equal to each other at final
  - use when only care about final difference between two groups

model2: answers if the two groups differ after adjusting for baseline difference
  - null hypothesis: the two group's means are equal to each other at final after adjusting for baseline
  - use when want to compare raw mean changes

model3: answers if the raw change from baseline to final differs between the two groups
  - use when want to present changes scores

model4: is the same as model2 except the coefficeient for Baseline is 1 less in model4
 - change = $\beta_0$ + $\beta_1$Baseline + $\beta_2$Group
 - Final - Baseline = $\beta_0$ + $\beta_1$Baseline + $\beta_2$Group 
 - Final = $\beta_0$ + $\beta_1$Baseline + Baseline + $\beta_2$Group 
 - Final = $\beta_0$ + ($\beta_1$ + 1)*Baseline + $\beta_2$Group 
 - We can see the only difference in model2 and model4 mathematically is the +1
 - use when baseline differences between the two groups and want to have outcome variable as changes scores

```{r}
lsmeans2 = lsmeans(model2, pairwise ~ group, adjust = "none")
summary(lsmeans2)
```


```{r}
model2_int = lm(Final ~ group * Baseline, data = df1_wide)
anova(model2, model2_int)
```