---
title: "Potato Randomization Analysis"
author: "Dina Dinh"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(easypackages)
library(psych)

libraries("readxl", "dplyr", "mmrm", "table1", "ggplot2", "lmerTest", "psych", "lsmeans", "knitr", "writexl", "car")
```

# Research Question: Did the intervention have any effect on LDL and HDL particle sizes?

```{r}
trt_randomization_df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/Potato randomization (5-21-21); Post-Study.xlsx")
trt_randomization_df2 = trt_randomization_df1 %>%
  rename(SubjectID = `Ppt. A#'s`,
         Comments = ...4) %>%
  filter(!is.na(SubjectID))

df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/Results from Labcorp with A numbers.xls")
colnames(df1) <- as.character(df1[4, ])
df2 = df1[-(1:4),]
colnames(df2)[3:4] <- c("SubjectID", "Visit")


df2 = df2 %>%
  mutate(#SubjectID = ifelse(SubjectID == "A00-079493", "A00-07-3493", SubjectID),
         SubjectID = ifelse(SubjectID == "A00-054384", "A00-05-4384", SubjectID),
         across(
      c(VLDLCP3, VLCP3, VMP3, VSP3, LDLP3, IDLP3, LLP3, LSP3, HDLP3, HLP3, HMP3, HSP3, VZ3, LZ3, HZ3, 
         LPIR, ELP_TC, ELP_HDLC, ELP_TG, ELP_APOB, ELP_LDLC, ELP_VLDLC, ELP_NONHDLC),
      ~ as.numeric(.)  
    ),
    Visit = as.factor(Visit))
```

## Data Cleaning and Summary Statistics

I believe subject "A00-079493" is a typo and should be "A00-07-3493" also "A00-054384" should be "A00-05-4384". However, since it is more evident for subject "A00-05-4384", I will only be adjusting this typo.

```{r}
df3 = df2 %>%
  dplyr::select(-Comment) %>%
  left_join(trt_randomization_df2, by = "SubjectID")
```

```{r}
demo_df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/PREMO Demographics (No PHI) - 2021-05-26 07 51 30.xlsx")
colnames(demo_df1) <- as.character(demo_df1[9, ])
demo_df2 = demo_df1[-(1:9),]

demo_df3 = demo_df2 %>%
  dplyr::select(-ScreeningNbr, -RandomizationNbr)
```

```{r}
df4 = demo_df3 %>%
  left_join(df3, by = "SubjectID")
```

```{r, eval = F}
df4 %>%
  group_by(SubjectID) %>%
  filter(n() == 1) %>%
  ungroup()

```

```{r}


df5 <- df4 %>%
  group_by(SubjectID, Visit) %>%
  filter(`Collection Date` == max(`Collection Date`, na.rm = TRUE)) %>%
  ungroup()

df5 %>%
  count(SubjectID, Visit) %>%
  filter(n > 1)

#write_xlsx(df5, "T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/v1_Potato_Randomization_Merged_Data_08212025.xlsx")

```

```{r} 

label(df3$VLDLCP3) = "VLDL & Chylomicron Particles (total)"
label(df3$VLCP3) = "Large VLDL & Chylomicron Particles"
label(df3$VMP3) = "Medium VLDL Particles"
label(df3$VSP3) = "Small VLDL Particles"
units(df3$VLDLCP3) = "nmol/L"
units(df3$VLCP3) = "nmol/L"
units(df3$VMP3) = "nmol/L"
units(df3$VSP3) = "nmol/L"

label(df3$LDLP3) = "LDL Particles (total)"
label(df3$IDLP3) = "IDL Particles"
label(df3$LLP3) = "Large LDL Particles"
label(df3$LSP3) = "Small LDL Particles (total)"
units(df3$LDLP3) = "nmol/L"
units(df3$IDLP3) = "nmol/L"
units(df3$LLP3) = "nmol/L"
units(df3$LSP3) = "nmol/L"

label(df3$HDLP3) = "HDL Particles (total)"
label(df3$HLP3) = "Large HDL Particles"
label(df3$HMP3) = "Medium HDL Particles"
label(df3$HSP3) = "Small HDL Particles"
units(df3$HDLP3) = "µmol/L"
units(df3$HLP3) = "µmol/L"
units(df3$HMP3) = "µmol/L"
units(df3$HSP3) = "µmol/L"

label(df3$VZ3) = "VLDL Size"
label(df3$LZ3) = "LDL Size"
label(df3$HZ3) = "HDL Size"
units(df3$VZ3) = "nm"
units(df3$LZ3) = "nm"
units(df3$HZ3) = "nm"

label(df3$LPIR) = "Lipoprotein Insulin Resistance Score"

label(df3$ELP_TC) = "Total Cholesterol"
label(df3$ELP_HDLC) = "HDL Cholesterol"
label(df3$ELP_TG) = "Triglycerides"
label(df3$ELP_APOB) = "ApoB"
label(df3$ELP_LDLC) = "LDL Cholesterol"
label(df3$ELP_VLDLC) = "VLDL Cholesterol"
label(df3$ELP_NONHDLC) = "Non HDL Cholesterol"
units(df3$ELP_TC) = "mg/dL"
units(df3$ELP_HDLC) = "mg/dL"
units(df3$ELP_TG) = "mg/dL"
units(df3$ELP_APOB) = "mg/dL"
units(df3$ELP_LDLC) = "mg/dL"
units(df3$ELP_VLDLC) = "mg/dL"
units(df3$ELP_NONHDLC) = "mg/dL"

table1(~ VLDLCP3 + VLCP3 + VMP3 + VSP3 + LDLP3 + IDLP3 + LLP3 + LSP3 + HDLP3 + HLP3 + HMP3 + HSP3 + VZ3 + LZ3 + HZ3 + 
         LPIR + ELP_TC + ELP_HDLC + ELP_TG + ELP_APOB + ELP_LDLC + ELP_VLDLC + ELP_NONHDLC | Diet * Visit, data = df5, overall = F)
```


Since our main focus is the LDL and HDL size changes in each diet groups over time, we'll plot those changes.

```{r}
summary_df3.1 <- df5 %>%
  group_by(Diet, Visit) %>%
  summarise(mean_LZ3 = mean(LZ3, na.rm = TRUE),
            se_LZ3 = sd(LZ3, na.rm = TRUE) / sqrt(sum(!is.na(LZ3))),
            .groups = "drop")

ggplot(summary_df3.1, aes(x = Visit, y = mean_LZ3, fill = Diet)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean_LZ3 - se_LZ3, ymax = mean_LZ3 + se_LZ3),
                position = position_dodge(width = 0.9),
                width = 0.1) +
  labs(
    title = "Average LDL Size over Timepoints",
    x = "Timepoint",
    y = "LDL Size (nm)"
  ) +
  theme_minimal() +
  coord_cartesian(ylim = c(20,21))
#mean(df3$LZ3[df3$Diet=="LMP" & df3$Visit == "VISIT 1"])
```

Even though the Average LDL size is smaller for the PLM diet group, the decrease in LDL size from Visit 1 to Visit 9 is almost the same for both groups. There are a lot of overalapping in the errorbars indicating that there is likely not a significant difference in the two diet groups. However, we still need to test these differences using linear mixed models.

```{r}
summary_df3.2 <- df5 %>%
  group_by(Diet, Visit) %>%
  summarise(mean_HZ3 = mean(HZ3, na.rm = TRUE),
            se_HZ3 = sd(HZ3, na.rm = TRUE) / sqrt(sum(!is.na(HZ3))),
            .groups = "drop")

ggplot(summary_df3.2, aes(x = Visit, y = mean_HZ3, fill = Diet)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean_HZ3 - se_HZ3, ymax = mean_HZ3 + se_HZ3),
                position = position_dodge(width = 0.9),
                width = 0.1) +
  labs(
    title = "Average HDL Size over Timepoints",
    x = "Timepoint",
    y = "HDL Size (nm)"
  ) +
  theme_minimal() +
  coord_cartesian(ylim = c(9,10))
```

We can see that the PLM diet group has larger HDL size, but the decrease in average HDL size from visit 1 to visit 9 is visibly more in the LMP group than the PLM. Again, there are strong overlapping error bars.


## Linear Mixed Models

Here, we will be using an unstructured covariance model. 

In running the model, I found that subjects "A00-01-2009" and "A00-07-8983" have two visit 1 observations recorded. I will use the most recent Visit 1 observation of those participants.


### LDL Size as the Response Variable

### Full Model

model1 = mmrm(LZ3 ~ Diet*Visit + us(Visit | SubjectID), data = df5)

```{r}

#model1 = lmer(LZ3 ~ 0 + Diet:Visit + (1 | SubjectID), data = df4)

model1 = mmrm(LZ3 ~ Diet*Visit + us(Visit | SubjectID), data = df5)
```

#### Model Summary

```{r}
paste0("AIC: ", AIC(model1))
summary(model1)
```
#### Type III Test

```{r}
Anova(model1, type = "3")
```

Here, the type III test is testing if 

1. $Diet_L$ = $Diet_P$

2. $Visit_1$ = $Visit_9$ 

3. ($Diet_9,_L$ - $Diet_1,_L$) = ($Diet_9,_P$ - $Diet_1,_P$)

These are possibles useful tests for the researchers.


```{r}
shapiro.test(residuals(model1))


summary_resid = psych::describe(residuals(model1))

summary_resid = summary_resid %>%
  dplyr::select(skew, kurtosis)
rownames(summary_resid) = NULL
```

```{r}
hist(residuals(model1), main = "Histogram of Residuals for LMM (LDL Size)", xlab = "residuals")
```

```{r}
lsmeans1 = lsmeans(model1, ~ Diet:Visit, adjust = "none")
lsmeans_summary = summary(lsmeans1, infer = c(TRUE, TRUE), adjust = "none")
#summary(lsmeans1)
```

```{r}
lsmeans_summary2 = lsmeans_summary %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)
```

```{r}
contrast1 <- list(
"Visit 1 Difference" =
c(-1,1,0,0)
)
contrast1.1 <- contrast(lsmeans1, method = contrast1, adjust = "none")

```

```{r}
contrast2 <- list(
"LMP: Visit 9 - Visit 1" =
c(-1,0,1,0)
)
contrast2.1 <- contrast(lsmeans1, method = contrast2, adjust = "none")
```

```{r}
contrast3 <- list(
"PLM: Visit 9 - Visit 1" =
c(0,-1,0,1)
)
contrast3.1 <- contrast(lsmeans1, method = contrast3, adjust = "none")
```

```{r}
contrast4 <- list(
"LMP vs PLM: Visit 9 - Visit 1" =
c(1,-1,-1,1)
)
contrast4.1 <- contrast(lsmeans1, method = contrast4, adjust = "none")
```

```{r}
contrast_df1 = as.data.frame(contrast1.1)
contrast_df2 = as.data.frame(contrast2.1)
contrast_df3 = as.data.frame(contrast3.1)
contrast_df4 = as.data.frame(contrast4.1)
```

```{r}
total_contrast_df1 = rbind(contrast_df1, contrast_df2, contrast_df3, contrast_df4)
```


#### Bar Charts 

```{r}
ggplot(lsmeans_summary2, aes(x = Visit, y = estimate, fill = Diet)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE),
                position = position_dodge(width = 0.9),
                width = 0.1) +
  labs(title = "Least Squares Means of LDL Size by Time",
       x = "Time",
       y = "LDL Size") +
  theme_minimal() +
  coord_cartesian(ylim = c(20,21))
```

#### Estimates

```{r}
kable(lsmeans_summary2)
```


#### Tests

```{r}
kable(total_contrast_df1)
```

### Reduced Model

model1.2 = mmrm(LZ3 ~ Diet:Visit + us(Visit | SubjectID), data = df5)

```{r}

#model1 = lmer(LZ3 ~ 0 + Diet:Visit + (1 | SubjectID), data = df4)

model1.2 = mmrm(LZ3 ~ 0 + Diet:Visit + us(Visit | SubjectID), data = df5)
```

#### Model Summary

```{r}
paste0("AIC: ", AIC(model1.2))
summary(model1.2)
```
#### Type III Test

```{r}
Anova(model1.2, type = "3")
```

Here, the type III test is testing if $\mu_1,_L$ = $\mu_1,_P$ = $\mu_9,_L$ = $\mu_9,_p$ = 0. **The numerator DF and Denominator DF is different from R than SAS and it's probably due to mmrm() DF  approximation algorithm.**

This is not a useful test is this case because LDL cannot be negative.

```{r}
shapiro.test(residuals(model1.2))

summary_resid.2 = psych::describe(residuals(model1.2))

summary_resid.2 = summary_resid.2 %>%
  dplyr::select(skew, kurtosis)
rownames(summary_resid.2) = NULL
```

```{r}
hist(residuals(model1.2), main = "Histogram of Residuals for LMM (LDL Size)", xlab = "residuals")
```

```{r}
lsmeans1.2 = lsmeans(model1.2, ~ Diet:Visit, adjust = "none")
lsmeans_summary.2 = summary(lsmeans1.2, infer = c(TRUE, TRUE), adjust = "none")
#summary(lsmeans1.2)
```

```{r}
lsmeans_summary2.2 = lsmeans_summary.2 %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)
```

```{r, eval=F}
contrast1.2 <- list(
  "Visit 1 Difference" =
    c(-1,1,0,0)
)
contrast1.1.2 <- contrast(lsmeans1.2, method = contrast1.2, adjust = "none")

```

```{r, eval=F}
contrast2.2 <- list(
  "LMP: Visit 9 - Visit 1" =
    c(-1,0,1,0)
)
contrast2.1.2 <- contrast(lsmeans1.2, method = contrast2.2, adjust = "none")
```

```{r, eval=F}
contrast3.2 <- list(
  "PLM: Visit 9 - Visit 1" =
    c(0,-1,0,1)
)
contrast3.1.2 <- contrast(lsmeans1.2, method = contrast3.2, adjust = "none")
```

```{r, eval=F}
contrast4.2 <- list(
  "LMP vs PLM: Visit 9 - Visit 1" =
    c(1,-1,-1,1)
)
contrast4.1.2 <- contrast(lsmeans1.2, method = contrast4.2, adjust = "none")
```

```{r, eval=F}
contrast_df1.2 = as.data.frame(contrast1.1.2)
contrast_df2.2 = as.data.frame(contrast2.1.2)
contrast_df3.2 = as.data.frame(contrast3.1.2)
contrast_df4.2 = as.data.frame(contrast4.1.2)
```

```{r, eval=F}
total_contrast_df1.2 = rbind(contrast_df1.2, contrast_df2.2, contrast_df3.2, contrast_df4.2)
```


#### Bar Charts 

```{r}
ggplot(lsmeans_summary2.2, aes(x = Visit, y = estimate, fill = Diet)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE),
                position = position_dodge(width = 0.9),
                width = 0.1) +
  labs(title = "Least Squares Means of LDL Size by Time",
       x = "Time",
       y = "LDL Size") +
  theme_minimal() +
  coord_cartesian(ylim = c(20,21))
```

#### Estimates

```{r}
kable(lsmeans_summary2.2)
```


#### Tests

```{r, eval=F}
kable(total_contrast_df1.2)
```


### HDL Size as the Response Variable

model1 = mmrm(HZ3 ~ Diet*Visit + us(Visit | SubjectID), data = df5)

```{r}

#model1 = lmer(HZ3 ~ 0 + Diet:Visit + (1 | SubjectID), data = df4)

model1 = mmrm(HZ3 ~ Diet*Visit + us(Visit | SubjectID), data = df5)
```

```{r}
shapiro.test(residuals(model1))


summary_resid = describe(residuals(model1))

summary_resid = summary_resid %>%
  dplyr::select(skew, kurtosis)
rownames(summary_resid) = NULL
```

```{r}
hist(residuals(model1), main = "Histogram of Residuals for LMM (HDL Size)", xlab = "residuals")
```

Although the Shapiro-Wilk test indicates that the residuals are not normally distributed, the histogram suggests that the residuals largely follow a normal distribution, with only a few outliers and is not heavily skewed. As such, the results of this analysis are expected to remain reliable despite the Shapiro-Wilk test's indication of non-normality.

```{r}
lsmeans1 = lsmeans(model1, ~ Diet:Visit, adjust = "none")
lsmeans_summary = summary(lsmeans1, infer = c(TRUE, TRUE), adjust = "none")
#summary(lsmeans1)
```

```{r}
lsmeans_summary2 = lsmeans_summary %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)
```

```{r}
contrast1 <- list(
"Visit 1 Difference" =
c(-1,1,0,0)
)
contrast1.1 <- contrast(lsmeans1, method = contrast1, adjust = "none")

```

```{r}
contrast2 <- list(
"LMP: Visit 9 - Visit 1" =
c(-1,0,1,0)
)
contrast2.1 <- contrast(lsmeans1, method = contrast2, adjust = "none")
```

```{r}
contrast3 <- list(
"PLM: Visit 9 - Visit 1" =
c(0,-1,0,1)
)
contrast3.1 <- contrast(lsmeans1, method = contrast3, adjust = "none")
```

```{r}
contrast4 <- list(
"LMP vs PLM: Visit 9 - Visit 1" =
c(1,-1,-1,1)
)
contrast4.1 <- contrast(lsmeans1, method = contrast4, adjust = "none")
```

```{r}
contrast_df1 = as.data.frame(contrast1.1)
contrast_df2 = as.data.frame(contrast2.1)
contrast_df3 = as.data.frame(contrast3.1)
contrast_df4 = as.data.frame(contrast4.1)
```

```{r}
total_contrast_df1 = rbind(contrast_df1, contrast_df2, contrast_df3, contrast_df4)
```


#### Bar Charts 

```{r}
ggplot(lsmeans_summary2, aes(x = Visit, y = estimate, fill = Diet)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE),
                position = position_dodge(width = 0.9),
                width = 0.1) +
  labs(title = "Least Squares Means of HDL Size by Time",
       x = "Time",
       y = "HDL Size") +
  theme_minimal() +
  coord_cartesian(ylim = c(9,10))
```

#### Estimates

```{r}
kable(lsmeans_summary2)
```


#### Tests

```{r}
kable(total_contrast_df1)
```


# 09/05/2025: Comparing Models

```{r}
model1 =  mmrm(HZ3 ~ Diet*Visit + us(Visit | SubjectID), data = df5)
model2 =  mmrm(HZ3 ~ Visit + Diet:Visit + us(Visit | SubjectID), data = df5)
model3 = mmrm(HZ3 ~ Diet + Diet:Visit + us(Visit | SubjectID), data = df5)
model4 = mmrm(HZ3 ~ 0 + Diet:Visit + us(Visit | SubjectID), data = df5)
```

## Comparing AICs

```{r}
aic_table = data.frame(
  Model= c("Full Model", "Visit Main Effect and Interaction", "Diet Main Effect and Interaction", "No intercept and Interaction"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4))
)

kable(aic_table)
```


## Comparing Model Summaries

### Full Model

```{r}
summary(model1)
```

### Visit Main Effect and Interaction

```{r}
summary(model2)
```


### Diet Main Effect and Interaction

```{r}
summary(model3)
```

### No intercept and Interaction

```{r}
summary(model4)
```

## Comparing Type III Tests

### Full Model

Here, the type III test is testing if 

1. $Diet_L$ = $Diet_P$

2. $Visit_1$ = $Visit_9$ 

3. ($Diet_9,_L$ - $Diet_1,_L$) = ($Diet_9,_P$ - $Diet_1,_P$)

These are possibles useful tests for the researchers.


```{r}
Anova(model1, type = "3")
```


### Visit Main Effect and Interaction

Here, the type III test is testing if 

1. $Visit_1$ = $Visit_9$ 

2. ???

```{r}
Anova(model2, type = "3")
```


### Diet Main Effect and Interaction

Here, the type III test is testing if 

1. $Diet_L$ = $Diet_P$

2. ???

```{r}
Anova(model3, type = "3")
```


### No Intercept and Interaction

Here, the type III test is testing if $\mu_1,_L$ = $\mu_1,_P$ = $\mu_9,_L$ = $\mu_9,_p$ = 0. **The numerator DF and Denominator DF is different from R than SAS and it's probably due to mmrm() DF  approximation algorithm.**

This is not a useful test is this case because HDL cannot be negative.

```{r}
Anova(model4, type = "3")
```

## Comparing LSMeans

### Full Model

```{r}
lsmeans1 = lsmeans(model1, ~ Diet:Visit, adjust = "none")
lsmeans_summary = summary(lsmeans1, infer = c(TRUE, TRUE), adjust = "none")
#summary(lsmeans1)
```

```{r}
lsmeans_summary2 = lsmeans_summary %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)

kable(lsmeans_summary2)
```


### Visit Main Effect and Interaction

```{r}
lsmeans1.2 = lsmeans(model2, ~ Diet:Visit, adjust = "none")
lsmeans_summary.2 = summary(lsmeans1.2, infer = c(TRUE, TRUE), adjust = "none")
```

```{r}
lsmeans_summary2.2 = lsmeans_summary.2 %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)

kable(lsmeans_summary2.2)
```

### Diet Main Effect and Interaction

```{r}
lsmeans1.3 = lsmeans(model3, ~ Diet:Visit, adjust = "none")
lsmeans_summary.3 = summary(lsmeans1.3, infer = c(TRUE, TRUE), adjust = "none")
```

```{r}
lsmeans_summary2.3 = lsmeans_summary.3 %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)

kable(lsmeans_summary2.3)
```


### No Intercept and Interaction

For some reason, the means of the PLM diet group is doubled here.

```{r}
lsmeans1.4 = lsmeans(model4, ~ Diet:Visit, adjust = "none")
lsmeans_summary.4 = summary(lsmeans1.4, infer = c(TRUE, TRUE), adjust = "none")
```

```{r}
lsmeans_summary2.4 = lsmeans_summary.4 %>%
  rename(estimate = lsmean) %>%
  dplyr::select(Diet, Visit, estimate, SE)

kable(lsmeans_summary2.4)
```


# What does it mean for the model to be the same?

The model is the same when you can get the same means from the model parameters and the relationship between predictor and response are the same for each model (units of predictors do not affect the models).


