---
title: "Potato Randomization Analysis"
author: "Dina Dinh"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(easypackages)
library(psych)

libraries("readxl", "dplyr", "mmrm", "table1", "ggplot2", "lmerTest", "psych", "lsmeans", "knitr", "writexl", "car")
```

# Research Question: Is the relationship between LDL and HDL the same between treatment groups?

```{r}
trt_randomization_df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/Potato randomization (5-21-21); Post-Study.xlsx")
trt_randomization_df2 = trt_randomization_df1 %>%
  rename(SubjectID = `Ppt. A#'s`,
         Comments = ...4) %>%
  filter(!is.na(SubjectID))

df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/Results from Labcorp with A numbers.xls")
colnames(df1) <- as.character(df1[4, ])
df2 = df1[-(1:4),]
colnames(df2)[3:4] <- c("SubjectID", "Visit")


df2 = df2 %>%
  mutate(#SubjectID = ifelse(SubjectID == "A00-079493", "A00-07-3493", SubjectID),
         SubjectID = ifelse(SubjectID == "A00-054384", "A00-05-4384", SubjectID),
         across(
      c(VLDLCP3, VLCP3, VMP3, VSP3, LDLP3, IDLP3, LLP3, LSP3, HDLP3, HLP3, HMP3, HSP3, VZ3, LZ3, HZ3, 
         LPIR, ELP_TC, ELP_HDLC, ELP_TG, ELP_APOB, ELP_LDLC, ELP_VLDLC, ELP_NONHDLC),
      ~ as.numeric(.)  
    ),
    Visit = as.factor(Visit))
```

## Data Cleaning and Summary Statistics

I believe subject "A00-079493" is a typo and should be "A00-07-3493" also "A00-054384" should be "A00-05-4384". However, since it is more evident for subject "A00-05-4384", I will only be adjusting this typo.

```{r}
df3 = df2 %>%
  dplyr::select(-Comment) %>%
  left_join(trt_randomization_df2, by = "SubjectID")
```

```{r}
demo_df1 = read_excel("T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/PREMO Demographics (No PHI) - 2021-05-26 07 51 30.xlsx")
colnames(demo_df1) <- as.character(demo_df1[9, ])
demo_df2 = demo_df1[-(1:9),]

demo_df3 = demo_df2 %>%
  dplyr::select(-ScreeningNbr, -RandomizationNbr)
```

```{r}
df4 = demo_df3 %>%
  left_join(df3, by = "SubjectID")
```

```{r, eval = F}
df4 %>%
  group_by(SubjectID) %>%
  filter(n() == 1) %>%
  ungroup()

```

```{r}


df5 <- df4 %>%
  group_by(SubjectID, Visit) %>%
  filter(`Collection Date` == max(`Collection Date`, na.rm = TRUE)) %>%
  ungroup()


#write_xlsx(df5, "T:/Biostatistics/Aa -- Dinh, Dina/Biostatistics Short Course/Potato Randomization/Data/v1_Potato_Randomization_Merged_Data_08212025.xlsx")

```

```{r} 

label(df5$VLDLCP3) = "VLDL & Chylomicron Particles (total)"
label(df5$VLCP3) = "Large VLDL & Chylomicron Particles"
label(df5$VMP3) = "Medium VLDL Particles"
label(df5$VSP3) = "Small VLDL Particles"
units(df5$VLDLCP3) = "nmol/L"
units(df5$VLCP3) = "nmol/L"
units(df5$VMP3) = "nmol/L"
units(df5$VSP3) = "nmol/L"

label(df5$LDLP3) = "LDL Particles (total)"
label(df5$IDLP3) = "IDL Particles"
label(df5$LLP3) = "Large LDL Particles"
label(df5$LSP3) = "Small LDL Particles (total)"
units(df5$LDLP3) = "nmol/L"
units(df5$IDLP3) = "nmol/L"
units(df5$LLP3) = "nmol/L"
units(df5$LSP3) = "nmol/L"

label(df5$HDLP3) = "HDL Particles (total)"
label(df5$HLP3) = "Large HDL Particles"
label(df5$HMP3) = "Medium HDL Particles"
label(df5$HSP3) = "Small HDL Particles"
units(df5$HDLP3) = "µmol/L"
units(df5$HLP3) = "µmol/L"
units(df5$HMP3) = "µmol/L"
units(df5$HSP3) = "µmol/L"

label(df5$VZ3) = "VLDL Size"
label(df5$LZ3) = "LDL Size"
label(df5$HZ3) = "HDL Size"
units(df5$VZ3) = "nm"
units(df5$LZ3) = "nm"
units(df5$HZ3) = "nm"

label(df5$LPIR) = "Lipoprotein Insulin Resistance Score"

label(df5$ELP_TC) = "Total Cholesterol"
label(df5$ELP_HDLC) = "HDL Cholesterol"
label(df5$ELP_TG) = "Triglycerides"
label(df5$ELP_APOB) = "ApoB"
label(df5$ELP_LDLC) = "LDL Cholesterol"
label(df5$ELP_VLDLC) = "VLDL Cholesterol"
label(df5$ELP_NONHDLC) = "Non HDL Cholesterol"
units(df5$ELP_TC) = "mg/dL"
units(df5$ELP_HDLC) = "mg/dL"
units(df5$ELP_TG) = "mg/dL"
units(df5$ELP_APOB) = "mg/dL"
units(df5$ELP_LDLC) = "mg/dL"
units(df5$ELP_VLDLC) = "mg/dL"
units(df5$ELP_NONHDLC) = "mg/dL"

table1(~ VLDLCP3 + VLCP3 + VMP3 + VSP3 + LDLP3 + IDLP3 + LLP3 + LSP3 + HDLP3 + HLP3 + HMP3 + HSP3 + VZ3 + LZ3 + HZ3 + 
         LPIR + ELP_TC + ELP_HDLC + ELP_TG + ELP_APOB + ELP_LDLC + ELP_VLDLC + ELP_NONHDLC | Diet * Visit, data = df5, overall = F)
```

```{r}
df6 = df5 %>%
  dplyr::select(1:12, Diet, HDLP3, LDLP3) %>%
  group_by(SubjectID) %>%
  mutate(change_HDL = ifelse(length(HDLP3[Visit == "VISIT 1"]) > 0 & length(HDLP3[Visit == "VISIT 9"]) > 0,
                        HDLP3[Visit == "VISIT 9"] - HDLP3[Visit == "VISIT 1"],
                        NA_real_),
         change_LDL = ifelse(length(LDLP3[Visit == "VISIT 1"]) > 0 & length(LDLP3[Visit == "VISIT 9"]) > 0,
                        LDLP3[Visit == "VISIT 9"] - LDLP3[Visit == "VISIT 1"],
                        NA_real_)) %>%
  ungroup() %>%
  filter(Visit == "VISIT 9")
#sum(df5$Visit == "VISIT 9")
```


Since our main focus is the LDL and HDL particle relationship between the diet groups, we'll plot those relationships.

```{r}
cor_test_l <- cor.test(df6$change_LDL[df6$Diet == "LMP"], df6$change_HDL[df6$Diet == "LMP"], method = "pearson")
    r_val_l <- round(cor_test_l$estimate, 3)
    p_val_l <- signif(cor_test_l$p.value, 3)
    
    # Build label
    label_l <- paste0("r = ", r_val_l, ", p-val = ", p_val_l)
    
cor_test_p <- cor.test(df6$change_LDL[df6$Diet == "PLM"], df6$change_HDL[df6$Diet == "PLM"], method = "pearson")
    r_val_p <- round(cor_test_p$estimate, 3)
    p_val_p <- signif(cor_test_p$p.value, 3)
    
    # Build label
    label_p <- paste0("r = ", r_val_p, ", p-val = ", p_val_p)
    
ggplot(df6, aes(x = change_HDL, y = change_LDL, color = Diet)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between change_LDL and change_HDL",
    subtitle = paste0("LMP: ", label_l, "\nPLM: ", label_p),
    x = "change_HDL",
    y = "change_LDL"
  ) +
  theme_minimal()
#mean(df3$LDLP3[df3$Diet=="LMP" & df3$Visit == "VISIT 1"])
```

```{r}
cor_test_l <- cor.test(df5$LDLP3[df5$Diet == "LMP"], df5$HDLP3[df5$Diet == "LMP"], method = "pearson")
    r_val_l <- round(cor_test_l$estimate, 3)
    p_val_l <- signif(cor_test_l$p.value, 3)
    
    # Build label
    label_l <- paste0("r = ", r_val_l, ", p-val = ", p_val_l)
    
cor_test_p <- cor.test(df5$LDLP3[df5$Diet == "PLM"], df5$HDLP3[df5$Diet == "PLM"], method = "pearson")
    r_val_p <- round(cor_test_p$estimate, 3)
    p_val_p <- signif(cor_test_p$p.value, 3)
    
    # Build label
    label_p <- paste0("r = ", r_val_p, ", p-val = ", p_val_p)
    
ggplot(df5, aes(x = HDLP3, y = LDLP3, color = Diet)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between LDL and HDL",
    subtitle = paste0("LMP: ", label_l, "\nPLM: ", label_p),
    x = "HDL",
    y = "LDL"
  ) +
  theme_minimal()
#mean(df3$LDLP3[df3$Diet=="LMP" & df3$Visit == "VISIT 1"])
```

There seems to be a relationship between LDL and HDL for the PLM diet group but not for the LMP group based on the correlation and p-value. Let's take a look at the correlation of each diet group for each Visit time. 

Again, the relationship between LDL and HDL is more noticeable for the PLM diet group versus LMP group.

```{r}
ggplot(df5, aes(x = HDLP3, y = LDLP3, color = Diet)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between LDL and HDL by Visit",
    x = "HDL",
    y = "LDL"
  ) +
  theme_minimal() +
  facet_wrap(~ Visit)
```


## Linear Mixed Models

Here, we will be using an unstructured covariance model. 

In running the model, I found that subjects "A00-01-2009" and "A00-07-8983" have two visit 1 observations recorded. I will use the most recent Visit 1 observation of those participants.


### Full Model

model1 = mmrm(LDLP3 ~ 0 + Diet:Visit + HDLP3:Diet:Visit + us(Visit | SubjectID), data = df5)

```{r}

#model1 = lmer(LZ3 ~ 0 + Diet:Visit + (1 | SubjectID), data = df5)

model1 = mmrm(LDLP3 ~ 0 + Diet:Visit + HDLP3:Diet:Visit + us(Visit | SubjectID), data = df5)
```


#### Model Summary

```{r}
paste0("AIC: ", AIC(model1))
summary(model1)
```



```{r}
model1_slope = data.frame(summary(model1)$coef)[-(1:4),]

model1_slope = model1_slope %>%
  rename(
    Std.Error = Std..Error,
    p.value = Pr...t..
  )
rownames(model1_slope) = c("LMP VIST 1", "PLM VIST 1", "LMP VIST 9", "PLM VIST 9")
```

```{r}
model1_int = data.frame(summary(model1)$coef)[(1:4),]

model1_int = model1_int %>%
  rename(
    Std.Error = Std..Error,
    p.value = Pr...t..
  )
rownames(model1_int) = c("LMP VIST 1", "PLM VIST 1", "LMP VIST 9", "PLM VIST 9")
```


```{r}
shapiro.test(residuals(model1))


summary_resid = psych::describe(residuals(model1))

summary_resid = summary_resid %>%
  dplyr::select(skew, kurtosis)
rownames(summary_resid) = NULL
```

```{r}
hist(residuals(model1), main = "Histogram of Residuals for LMM", xlab = "residuals")
```

### Type III Test

```{r}
Anova(model1, type = "3")
```

Here, the type III test is testing if 

1. $Diet_L,_9$ = $Diet_P,_9$ = $Diet_L,_1$ = $Diet_P,_1$ = 0 (Intercepts)

2. $Diet_L,_9$ = $Diet_P,_9$ = $Diet_L,_1$ = $Diet_P,_1$ = 0 (Slopes)


```{r, eval=FALSE}
#======Testing Type III Test Results===============


# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0.25, 0.25, -0.25, -0.25, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

type3_test1 <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
```


```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0.5, 0, 0.5, 0, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

lmp_int1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(lmp_int1_table) <- "LMP"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0.5, 0, 0.5, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

plm_int1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(plm_int1_table) <- "PLM"
```

```{r}
model1_int_summary = rbind(lmp_int1_table, plm_int1_table, model1_int)

model1_int_summary2 = model1_int_summary %>%
  dplyr::select(-df, -t.value, -p.value)
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(-1, 0, 1, 0, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test5.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test5.1_base_table) <- "LMP: VIST 9-1 Intercept"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, -1, 0, 1, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test6.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test6.1_base_table) <- "PLM: VIST 9-1 Intercept"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(1, -1, 0, 0, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test1.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test1.1_base_table) <- "LMP - PLM : VIST 1 Intercept"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 1, -1, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test2.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test2.1_base_table) <- "LMP - PLM : VIST 9 Intercept"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(1, -1, -1, 1, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test4.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test4.1_base_table) <- "LMP - PLM : VIST 9-1 Intercept"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0.5, 0.5, -0.5, -0.5, 0, 0, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test3.1_base_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test3.1_base_table) <- "LMP - PLM : Overall"
```

```{r}
test_base_result1 = rbind(test3.1_base_table, test5.1_base_table, test6.1_base_table, test1.1_base_table, test2.1_base_table, test4.1_base_table)
```


```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 0.5, 0, 0.5, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

lmp_slope1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(lmp_slope1_table) <- "LMP"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 0, 0.5, 0, 0.5)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

plm_slope1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(plm_slope1_table) <- "PLM"
```

```{r}
model1_slope_summary = rbind(lmp_slope1_table, plm_slope1_table, model1_slope)
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, -1, 0, 1, 0) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test5.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test5.1_table) <- "LMP : VISIT 9-1 Slope"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 0, -1, 0, 1) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test6.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test6.1_table) <- "PLM : VISIT 9-1 Slope"
```


```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 1, -1, 0, 0) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test1.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test1.1_table) <- "LMP - PLM : VISIT 1 Slope"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 0, 0, 1, -1) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test2.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test2.1_table) <- "LMP - PLM : VISIT 9 Slope"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, 1, -1, -1, 1) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test4.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test4.1_table) <- "LMP - PLM : VISIT 9-1 Slope"
```

```{r}
# Extract coefficient estimates
coef_estimates <- coef(model1)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model1)

# Define contrast vector
K <- c(0, 0, 0, 0, .5, .5, -.5, -.5) 

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df5$SubjectID))  # Number of unique subjects
p <- length(coef(model1))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

test3.1_table <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
rownames(test3.1_table) <- "LMP - PLM : Overall"
```


```{r}
test_result1 = rbind(test3.1_table, test5.1_table, test6.1_table, test1.1_table, test2.1_table, test4.1_table)
```

### Intercept Estimates

```{r}
kable(model1_int_summary2)
```


### Tests for Intercept

All contrasts are in the format of LMP - PLM.

```{r}
kable(test_base_result1)
```

### Slope Estimates

```{r}
kable(model1_slope_summary)
```


### Tests for Slope

All contrasts are in the format of LMP - PLM.

```{r}
kable(test_result1)
```

## Using Change Scores

### No Intercept Model

```{r}
model2 = glm(change_LDL ~ 0 + Diet + change_HDL:Diet, data = df6)
```


```{r}
summary(model2)
```

```{r}
Anova(model2, type = "3")
```

Here, the type III test is testing if 

1. $Diet_L$ = $Diet_P$ = 0 

2. $Diet_L$:change_HDL = $Diet_P$:change_HDL = 0


Can't add things that don't have the same units. Contrasts tests if things equal to 0. 1, -1 tests if the difference is equal 0.
When Num DF is > 1, then you can't do a contrast. Contrast only have 1 equal sign.

```{r}

# Extract coefficient estimates
coef_estimates <- coef(model2)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model2)

# Define contrast vector
#K <- c(1, -1, 0, 0)
K <- c(0, 0, 1, -1)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df6$SubjectID))  # Number of unique subjects
p <- length(coef(model2))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df) # pt(-abs(t_stat), df) 

slope_test <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
```

```{r}

# Extract coefficient estimates
coef_estimates <- coef(model2)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model2)

# Define contrast vector
#K <- c(1, -1, 0, 0)
K <- c(1, -1, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df6$SubjectID))  # Number of unique subjects
p <- length(coef(model2))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df) # pt(-abs(t_stat), df) 

int_test <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
```

#### Intercept Test

```{r}
kable(int_test)
```

#### Slope Test

```{r}
kable(slope_test)
```

### Full Model

```{r}
model3 = glm(change_LDL ~ change_HDL*Diet, data = df6)
```


```{r}
summary(model3)
```

```{r}
Anova(model3, type = "3")
```

Here, the type III test is testing if 

1. change_HDL = 0 : K <- c(0, 1, 0, 0) *Not testing the same thing as SAS bc the p-val is diff.

2. $Diet_L$ - $Diet_P$ = 0 (Intercepts of the lines)

3. $Diet_L$:change_HDL - $Diet_P$:change_HDL = 0

```{r, eval=FALSE}
#======Testing Type III Test Results===============


# Extract coefficient estimates
coef_estimates <- coef(model3)

# Extract variance-covariance matrix for fixed effects
vcov_matrix <- vcov(model3)

# Define contrast vector
K <- c(0, 1, 0, 0)

# Calculate the estimate for the contrast
estimate <- sum(K * coef_estimates)

# Calculate the variance of the contrast
var_contrast <- as.numeric(t(K) %*% vcov_matrix %*% K)
std_error <- sqrt(var_contrast)

# Calculate t-statistic
t_stat <- estimate / std_error

n_subjects <- length(unique(df6$SubjectID))  # Number of unique subjects
p <- length(coef(model3))  # Number of fixed effects

# Approximate degrees of freedom
df <- n_subjects - p

# Calculate two-tailed p-value
p_value <- 2 * pt(-abs(t_stat), df)

type3_test2 <- data.frame(
  Estimate = estimate,
  Std.Error = std_error,
  df = df,
  t.value = t_stat,
  p.value = p_value
)
```
