---
title: "Correlation and Slope Comparison"
author: "Dina Dinh"
date: "2025-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r}
library(easypackages)
libraries("dplyr", "tidyr", "ggplot2")
```

# Comparing P-Values

```{r}
set.seed(1)
N = 100

# 1.
d1 = tibble(
  x = runif(N, 0, 1),
  y = sin(0.5 * pi * x) + rnorm(N, 0, 3)
)

# 2.
d2 = tibble(
  x = runif(N, -2, 2),
  y = 1.2 * x + rnorm(N, 0, 0.5)
)

# 3.
d3 = tibble(
  x = c(rnorm(N - 2, 0, 1), 6, 7),
  y = c(rnorm(N - 2, 0, 1), 10, 14)
)

# 4.
d4 = tibble(
  x = rnorm(N),
  y = rnorm(N)
)
```

## Case 1

```{r}
cor_test <- cor.test(d1$x, d1$y, method = "pearson")
    r_val <- round(cor_test$estimate, 3)
    p_val <- signif(cor_test$p.value, 3)
    #p_val = ifelse(p_val < 0.001, "p-val < 0.001", paste0("p-val = ", signif(p_val, 3)))
    
    # Build label
    label <- paste0("r = ", r_val, ", p-val = ", p_val)
    
fit <- lm(y ~ x, data = d1)
pval_lm <- paste0("slope p-val = ", signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3))

#pval_lm <- ifelse(pval_lm < 0.001, "slope p-val < 0.001", paste0("slope p-val = ", signif(pval_lm, 3)))

ggplot(d1, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = pval_lm,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Case 2

```{r}
cor_test <- cor.test(d2$x, d2$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)
#p_val = ifelse(p_val < 0.001, "p-val < 0.001", paste0("p-val = ", signif(p_val, 3)))

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = d2)
pval_lm <- paste0("slope p-val = ", signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3))

#pval_lm <- ifelse(pval_lm < 0.001, "slope p-val < 0.001", paste0("slope p-val = ", signif(pval_lm, 3)))

ggplot(d2, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = pval_lm,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Case 3

```{r}
cor_test <- cor.test(d3$x, d3$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)
#p_val = ifelse(p_val < 0.001, "p-val < 0.001", paste0("p-val = ", signif(p_val, 3)))

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = d3)
pval_lm <- paste0("slope p-val = ", signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3))

#pval_lm <- ifelse(pval_lm < 0.001, "slope p-val < 0.001", paste0("slope p-val = ", signif(pval_lm, 3)))

ggplot(d3, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = pval_lm,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Case 4

```{r}
cor_test <- cor.test(d4$x, d4$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)
#p_val = ifelse(p_val < 0.001, "p-val < 0.001", paste0("p-val = ", signif(p_val, 3)))

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = d4)
pval_lm <- paste0("slope p-val = ", signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3))

#pval_lm <- ifelse(pval_lm < 0.001, "slope p-val < 0.001", paste0("slope p-val = ", signif(pval_lm, 3)))

ggplot(d4, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = pval_lm,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Result

They produce the exact same p-value because both are testing whether a linear association exists between x and y. The correlation p-val would produce different p-val as slope if we were using a non-linear association test like Spearman.


# Comparing Correlation and Slope Values

```{r}
set.seed(1)

N <- 100
x <- scale(rnorm(N), center = TRUE, scale = TRUE)  # centered & scaled for stability
b_target <- 1.5                                     # target slope for A1/A2

# Build a noise vector orthogonal to x so OLS slope is EXACTLY b_target
z <- rnorm(N)
u <- as.numeric(z - x * (sum(x * z) / sum(x^2)))   # u ⟂ x  (x'u = 0)
# sanity check: crossprod(x, u) ~ 0

# ------- Part A: SAME slope, DIFFERENT correlations -------
# y = b*x + k*u  => OLS slope = b for ANY k because x'u = 0
k1 <- 0.8
k2 <- 4.0
y_A1 <- b_target * x + k1 * u
y_A2 <- b_target * x + k2 * u

A1 <- data.frame(x = as.numeric(x), y = as.numeric(y_A1))
A2 <- data.frame(x = as.numeric(x), y = as.numeric(y_A2))

# ------- Part B: SAME correlation, DIFFERENT slopes -------
# Scaling y by a positive constant changes slope but NOT correlation
# Start from a clean linear dataset
y_base <- 0.9 * x + rnorm(N, 0, 0.6)
B1 <- data.frame(x = as.numeric(x), y = as.numeric(y_base))          # slope ~ 0.9
B2 <- data.frame(x = as.numeric(x), y = as.numeric(3 * y_base))      # slope ~ 2.7
# cor(B1$x, B1$y) == cor(B2$x, B2$y) in the sample (up to rounding)

```

## Same Slopes, Different Correlations

```{r}
cor_test <- cor.test(A1$x, A1$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = A1)
pval_lm <- signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3)
slope_lm <- signif(summary(fit)$coefficients["x", "Estimate"], 3)

label2 <- paste0("slope = ", slope_lm, ", p-val = ", pval_lm)

ggplot(A1, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = label2,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

```{r}
cor_test <- cor.test(A2$x, A2$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = A2)
pval_lm <- signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3)
slope_lm <- signif(summary(fit)$coefficients["x", "Estimate"], 3)

label2 <- paste0("slope = ", slope_lm, ", p-val = ", pval_lm)

ggplot(A2, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = label2,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Same Correlations, Different Slopes

```{r}
cor_test <- cor.test(B1$x, B1$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = B1)
pval_lm <- signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3)
slope_lm <- signif(summary(fit)$coefficients["x", "Estimate"], 3)

label2 <- paste0("slope = ", slope_lm, ", p-val = ", pval_lm)

ggplot(B1, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = label2,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

```{r}
cor_test <- cor.test(B2$x, B2$y, method = "pearson")
r_val <- round(cor_test$estimate, 3)
p_val <- signif(cor_test$p.value, 3)

# Build label
label <- paste0("r = ", r_val, ", p-val = ", p_val)

fit <- lm(y ~ x, data = B2)
pval_lm <- signif(summary(fit)$coefficients["x", "Pr(>|t|)"], 3)
slope_lm <- signif(summary(fit)$coefficients["x", "Estimate"], 3)

label2 <- paste0("slope = ", slope_lm, ", p-val = ", pval_lm)

ggplot(B2, aes(x = x, y = y)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between x and y",
    subtitle = paste0("Corr: ", label),
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,           
    hjust = 1.1, vjust = -0.5,
    label = label2,
    size = 4,
    color = "blue"
  ) +
  theme_minimal()
```

## Result

We can see that the p-val for the correlation and slopes are the same in each case; however the values of the slopes and correlations themselves can change while the other stays constant.

